---
title: "BamBamBam"
author: "kailey wolfe"
date: "11/8/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##check data types
## dont use id
## x_cols = c('incident_id','date', etc)

##caret zv, nzv in preProcess
##dmv = create DummyVars
##predict(dmv, data)

##find outliers - create linear model - fit(residuals)
#filter data through fit(residuals)

#####info USE SPLIT ON || 
#regex

#lubridate -> create c('month','year')
```
##Are gun violence incidents increasing in the United States? How many predicted deaths in December of 2018?

##load data
```{r}
#file path
file_path = "bambambam/gun-violence-data.csv"
d = read.csv(file_path)
#View(d)

cols = colnames(d)
cols

##add missing row
missing = data.frame('na', '10/1/17', 'Nevada', 'Las Vegas', 59, 489, 'Mandalay Bay 3950 Blvd S','-','-','Route 91 Harvest Festiva; concert, open fire from 32nd floor. 47 guns seized; TOTAL:59 kill, 489 inj, number shot TBD,girlfriend Marilou Danley POI', '-','Hotel','-',47,'64', '-', '-', '-', '-', '-','-')
names(missing) = cols

df = rbind(d, missing)
```
## add year, month, day, total loss
```{r}
library(lubridate)
date = df$date
strD = as.character(date)
dates = as.Date(strD, "%m/%d/%y")

gun = df
gun$date = dates

year = year(dates)
month = month(dates)
day = day(dates)
weekday = wday(dates, label=TRUE)

gun$year = year
gun$month = month
gun$day = day
gun$weekday = weekday
gun$total_loss = gun$n_killed + gun$n_injured
#summary(gun)
```

##find and plot the number of actual gun violence incidents per year (not including 2018 or 2013)
```{r}
library(ggplot2)
ag = aggregate(gun$n_killed, by=list(gun$year), FUN=sum)
ag = ag[-c(1,6), ]
colnames(ag) = c('Year','total')

p = ggplot(ag)
p = p + geom_bar(mapping=aes(x=ag$Year, y=ag$total), stat='identity', color='red', fill='red', width = .25) + xlab("Year") + ylab("total killed") +ggtitle("Number of Deaths from 2014 - 2017") + theme_minimal()
p
```
## total killed / injuries
```{r}
library(ggplot2)
#x = gun$year
#y = total number of deaths
ag2 = aggregate(gun$total_loss, by=list(gun$year), FUN=sum)
ag2 = ag2[-c(1,6), ]
colnames(ag2) = c('Year','total')

p2 = ggplot(ag2)
p2 = p2 + geom_bar(mapping=aes(x=ag2$Year, y=ag2$total), stat='identity',color='orange', fill='orange', width = .25) + xlab("Year") + ylab("total killed or injured") +ggtitle("Number of Deaths and Injuries from 2014 - 2017") + theme_minimal()
p2
```

### all in one graph 
```{r}
library(ggplot2)

ag$incidents = "number of fatalities"
ag2$incidents = "number of fatalities and injuries"
d = rbind(ag, ag2)
bigplot <- ggplot(d, aes(d$Year, d$total, fill=incidents)) + geom_bar(stat="identity", position=position_dodge()) + xlab("Year") + ylab("Total") +ggtitle("Fatalities and Injuries 2014 - 2017") + theme_minimal()
bigplot
```
##plot number of incidents from 2014 - 2017
```{r}
library(ggplot2)
library(plyr)

incidents = count(year, "year")
incidents = incidents[-c(1,6), ]

p3 = ggplot(incidents)
p3 = p3 + geom_bar(mapping=aes(x=incidents$year, y=incidents$freq), stat='identity',color='blue', fill='blue', width = .25) + xlab("Year") + ylab("Gun Violence Incidents") +ggtitle("Gun Violence Incidents from 2014 - 2017") + theme_minimal()
p3
```
##predict the total number of incidents, deaths, total deaths + injuries for 2018 and 2019
```{r}
library(lattice)
library(ggplot2)
library(caret)

ag = aggregate(gun$n_killed, by=list(gun$year), FUN=sum)
ag = ag[-c(1,6), ]
colnames(ag) = c('Year','total')
fit = lm(total~Year, ag)

fit$coefficients
plot(ag$Year,ag$total)
abline(fit$coefficients, col='red')

summary(fit)
```

```{r}
total_18 = data.frame(2018,predict(fit, data.frame(Year=2018)))
total_19 = data.frame(2019,predict(fit, data.frame(Year=2019)))

names(total_18)=c('Year','total')
names(total_19)=c('Year','total')
modified_ag = rbind(ag,total_18)
mod_mod_ag = rbind(modified_ag, total_19)

fit = lm(total~Year, mod_mod_ag)
#fit$coefficients

plot(mod_mod_ag$Year,mod_mod_ag$total)
abline(fit$coefficients, col='red')
```
